---
title: "data_607_project_2_dataset_1"
author: "Maxfield Raynolds"
date: "2025-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
```


# Project 2 - Dataset 1

## Tidying

### Load the Dataset
```{r load the dataset}
data1 <- read.csv("https://raw.githubusercontent.com/mraynolds/data_607/refs/heads/main/data_607_project_2_dataset_1.csv")

glimpse(data1)
```
### Pivot data longer

The dataset has one primary issue keeping it from being tidy. The header columns are combinations of variable, both a month and an environmental measurement (temperature and humidity).

The following code pivots the data longer while separating the column names into their separate components.

```{r pivot longer and separate header roles}
data1 <- data1 |> 
  pivot_longer(
    cols = !City,
    names_to = c("environment", "month"),
    names_sep = "_",
    values_to = "measurement"
  )

glimpse(data1)
```
### Create a factor for months

```{r factor for months}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```


### Clean up the data

The data now could use some cleaning.

The code block below does the following:

- renames "Humid" to "humidity_%" to indicate that the measurement is "percent humidity"
- renames "Temp" to "temp_f" to indicate that the measurement is "Temperature in Fahrenheit"
- removes all unit measurements from the measurement column (degrees Fahrenheit and percent symbol) so that the digits can be used as numbers
- converts the measurement column to numeric

```{r mutate all columns}
data_long <- data1 |>
  mutate(
    environment = str_replace_all(environment,"Humid","humidity_pct"),
    environment = str_replace_all(environment,"Temp","temp_f"),
    measurement = parse_number(measurement),
    measurement = as.numeric(measurement),
    month = factor(month, levels = month_levels),
    city = City
  )

glimpse(data_long)
```
### Pivot Wider

The data in the previous data frame, "data_long", is still not tidy as there are two types of data in the measurement column, humidity and temperature.

The following code block pivots the data wider so that temperature and humidity are separated into their own columns.

```{r pivot wider}
data_tidy <- data_long |> 
  pivot_wider(
    names_from = environment,
    values_from = measurement
  )

glimpse(data_tidy)
```

## Analysis

# Averages

The code block below calculates the average temperature and humidity for the dataset, first by month, and then by city.

```{r average metrics by month and city}
data_tidy |>
  group_by(month) |> 
  summarise(
    avg_temp_f = round(mean(temp_f),1),
    avg_humidity_pct = round(mean(humidity_pct),1)
  )

data_tidy |>
  group_by(City) |> 
  summarise(
    avg_temp_f = round(mean(temp_f),1),
    avg_humidity_pct = round(mean(humidity_pct),1)
  )
```

# Plot

A plot of temperature over time.

```{r plot of temp}
ggplot(data_tidy, aes(x = month, y = temp_f, group = city, color = city)) + 
  geom_line()
```
A plot of humidty over time.

```{r plot of humidity}
ggplot(data_tidy, aes(x = month, y = humidity_pct, group = city, color = city)) + 
  geom_line()
```
# Plot of temp and humidity together

```{r plot of temp and humidity}
ggplot(data_tidy, aes(x = month)) + geom_line(aes(y = humidity_pct, group = city, color = city)) +
  geom_line(aes(y = temp_f, group = city, color = city)) +
  scale_y_continuous(name = "Percent Humidity", sec.axis = sec_axis(~., name = "Temperature in Fahrenheit")) 
```